Practical Optional Types for Clojure

### 2015-02-27 overview.tex
describe nonreflective syntax
overview, Thursday evening changes

@@

We introduce Clojure's syntax.
Parenthesis indicate \emph{applications}, square brackets
are syntax for literal \emph{vectors} and curly braces delimit literal
\emph{hash-maps}.
\clj{nil} is exactly Java's \java{null} and is one of two logically
false values (the other is \clj{false}).
\emph{Symbols} begin with an alphabetic character.
A symbol prefixed with a colon like \clj{:a} is an interned \emph{keyword}.
The \clj{get} function takes a hash-map and a key returns the value
associated with the key in the map, or returns \clj{nil} if there is none.
Keywords are functions that take a map and applies \clj{get} to the map
and itself.
\clj{inc} and \clj{dec} increment and decrement numbers respectively.
\clj{assoc} and \clj{dissoc} associate and dissociate entries
from immutable maps.
The \clj{=} function returns true if all its arguments are structurally equal---in this
paper we use it only for pointer equality.
Everything is an instance of some class
except for \clj{nil}. Predicates like \clj{number?} are equivalent to
a Java \java{instanceof} test, in this case on \clj{java.lang.Number}.

Clojure uses \emph{namespaces}, declared with the
\clj{ns} form, to manage top-level \emph{vars} and
class imports.
Example~\ref{example:conditionalflow}
is a regular Clojure file compiled with
the Clojure compiler using \clj{:require} to declare a
runtime dependency on Typed Clojure's core namespace
\clj{clojure.core.typed}.
Programmers import typed versions of certain constructs as needed, like
a \clj{fn} variant that supports annotations,
and use the provided \clj{check-ns} function to type check the current namespace
when convenient.
We omit \clj{ns} forms for the rest of the examples.
\footnote{See <this> github repo for full working examples}

\emph{Destructuring} is binding position syntax for pattern matching.
A \emph{map destructuring} contains pairs of symbols and keywords that bind the symbols
to lookups on that keyword with the current argument, optionally terminated by
\clj{:as} and a symbol, which binds the symbol to the current argument. For example
$$
\clj{(fn [{^File x :x, :as m}] ...)}
$$ 
expands to a \clj{let} that binds \clj{m} to the argument and 
\clj{^File x} to \clj{(:x m)}.
Destructuring is supported similarly by \clj{let}.

@@

Along with conditional control flow,
Clojure programmers rely on \emph{exceptions}
to assert type-related invariants.

@@

To check this example, we note that
occurrence typing already provides
valuable information about the conditional expression---namely that it cannot return logically true and if it returns logically false 
then \clj{x} is a \clj{Number}. Equivalently, if it returns a value then \clj{x} is a \clj{Number}.
We use this assumption to check \clj{(inc x)} with an automatic
guarantee that a null-pointer exception is impossible
(modelled formally in section~\ref{sec:doformal} and proved
in section~\ref{sec:metatheory}).

@@

\begin{Code}
\begin{exmp}
\inputminted[firstline=6,lastline=13]{clojure}{code/demo/src/demo/hmap.clj}
\label{example:decleaf}
\end{exmp}
\end{Code}

Here we define a type abbreviation with \clj{defalias} called \clj{Expr}
that describes the structure of a recursively-defined AST as a union of HMaps.
\clj{an-exp} is a function verified to return an \clj{Expr}.

The next example is a real Typed Clojure program written by CircleCI, who maintain a large production
installation of Clojure (see section~\ref{sec:casestudy} for a case study).

\begin{Code}
\begin{exmp}
\inputminted[firstline=10,lastline=22]{clojure}{code/demo/src/demo/key.clj}
\end{exmp}
\end{Code}

\clj{enc-keypair} takes an unencrypted keypair and returns an encrypted keypair by
removing the raw \clj{:private-key} entry and associating an encrypted private key
as \clj{:enc-private-key}.
Since \clj{EncKeyPair} is \clj{:complete?}, Typed Clojure enforces the return type
does not contain an entry \clj{:private-key}, and would complain if the \clj{dissoc}
operation forgot to remove it.

\begin{Code}
\begin{exmp}
\inputminted[firstline=10,lastline=23]{clojure}{code/demo/src/demo/key2.clj}
\end{exmp}
\end{Code}

HMaps interact with occurrence typing in interesting ways.

\begin{Code}
\begin{exmp}
\inputminted[firstline=15,lastline=27]{clojure}{code/demo/src/demo/hmap.clj}
\end{exmp}
\end{Code}

@@

Clojure supports interoperability with Java, including the ability to
call constructors, methods and access fields.
Method calls use prefix syntax, for example this call to 
\clj{java.io.File}'s \clj{getParent} method.

@@

We have a problem: Clojure's compiler does not know which method to call.
Instead, \clj{f} is inspected at runtime and an appropriate method is selected based
and then invoked by \emph{Java reflection}.
Unfortunately, reflection is slow and Clojure's algorithm for
choosing methods at runtime is undefined in some cases.

To resolve reflective calls, Clojure supports \emph{type hints}.

@@

The Clojure compiler uses the type hint on \clj{f}
to statically resolve the method call to the \clj{java.io.File}
method with the Java signature

\begin{minted}{java}
  public String getParent()
\end{minted}

Typed Clojure ignores type hints, so to check this we must
annotate the parameter's static type.

\begin{minted}{clj}
  (fn [f :- File] (.getParent f))
\end{minted}

However Typed Clojure disallows reflection in typed code, so we
must add a type hint.

@@

Now we have a well-typed expression free of reflection.

@@

\begin{exmp}
\inputminted[firstline=12,lastline=13]{clojure}{code/demo/src/demo/parent3.clj}
\end{exmp}

@@

\begin{Code}
\begin{exmp}
\inputminted[firstline=15,lastline=16]{clojure}{code/demo/src/demo/parent3.clj}
\end{exmp}
\end{Code}

Notice a type hint is used in an argument position to help choose the \java{File(String pathname)}
constructor. By default Typed Clojure conservatively assumes method and constructor arguments to be \emph{non-nullable},
but can be configured globally for particular positions if needed.

Finally, Typed Clojure guarantees typed code cannot throw a null-pointer exception
by attempting to dereference \clj{nil}. This can happen in several ways, including
calling a method or a field on \clj{nil}.
The next example calls a method on once-nullable local binding.
(\clj{ann} annotates a var with an expected type).

\begin{exmp}
\inputminted[firstline=5,lastline=8]{clojure}{code/demo/src/demo/parent3.clj}
\end{exmp}

The test on \clj{f} is essential to check this example---only then can Typed Clojure
prove null-pointer exceptions are impossible.

@@

A multimethod in Clojure is a function that contains a dispatch
function and methods. Multimethods are created with {\clj{defmulti}}.
\begin{minted}{clojure}
(ann rep [Any -> String])
(defmulti rep class)
\end{minted}
\clj{rep} is a multimethod of type \clj{[Any -> String]} with an empty \emph{dispatch table}
and \emph{dispatch function} \clj{class}, a function that returns the argument's class or \clj{nil} if none.
There are no methods to dispatch to so invocations of \clj{rep} will fail at runtime.
\begin{minted}{clojure}
  (rep :a) ;=> IllegalArgumentException ...
\end{minted}

Methods are installed via {\clj{defmethod}}.
\begin{minted}{clojure}
(defmethod rep Keyword [x] (str (name x)))
\end{minted}
We extend \clj{rep}'s dispatch table, mapping
the \emph{dispatch value} \clj{Keyword} to the function
\clj{(fn [x] (str (name x)))}. The call \clj{(rep arg)}
uses the value of \clj{(isa? (disp-fn arg) disp-val)}
on each method entry to pick a winner \clj{mth}, and finally returns \clj{(mth arg)}.
\clj{isa?} is a subclassing check when provided with classes
\begin{minted}{clojure}
  (isa? Keyword Object) ;=> true
  (isa? Keyword Number) ;=> false
\end{minted}
otherwise an equality check.
\begin{minted}{clojure}
  (isa? :a :a) ;=> true
  (isa? :a 1) ;=> false
\end{minted}

For example
\clj{(rep :a)}
picks the \clj{Keyword} method because
\clj{(isa? (class :a) Keyword)} returns true,
and finally returns \clj{"a"}, the value of
\clj{((fn [x] (str (name x))) :a)}.
We give the full definition of \clj{rep}.

\begin{Code}
\begin{exmp}
\inputminted[firstline=5,lastline=11]{clojure}{code/demo/src/demo/rep.clj}
\label{example:rep}
\end{exmp}
\end{Code}
Typed Clojure does not statically ensure multimethod calls dispatch successfully---
\clj{(rep "a")} type checks but throws a runtime error.

The flexibility of \clj{isa?} is key to the generality of multimethods. 
We can dispatch on the \clj{:op} key 
of our HMap AST \clj{Expr}.
Keywords are functions that look themselves up in their argument, so \clj{:op}
is our dispatch function.
\begin{Code}
\begin{exmp}
\inputminted[firstline=5,lastline=25]{clojure}{code/demo/src/demo/eg5.clj}
\end{exmp}
\end{Code}
The destructuring syntax \clj{(fn [{:keys [val] :as m}] ...)} binds
\clj{m} to the value of the first argument, and \clj{val} to \clj{(:val m)}.

\clj{isa?} is special with vectors---vectors of the
same length recursively call \clj{isa?} on the elements pairwise.
\begin{minted}{clojure}
  (isa? [Keyword :a] [Object :a]) ;=> true
  (isa? [Keyword Keyword] [Object Object]) ;=> true
  (isa? [Keyword Object] [Object Object]) ;=> false
\end{minted}

Now multiple dispatch is possible---we dispatch on the class of both
arguments simultaneously by defining a dispatch function that returns
a vector containing the classes of the arguments.
\begin{Code}
\begin{exmp}
\inputminted[firstline=6,lastline=23]{clojure}{code/demo/src/demo/eg7.clj}
\end{exmp}
\end{Code}
The dispatch values are also vectors---the first method is picked
when the left argument is a \clj{Number} and the right is a \clj{Keyword},
and is thus safe to increment and extract its name respectively.
The dispatch value \clj{:default} specifies a default
method if no preferred method is found.

No extra annotations are needed to follow type-directed control flow
in multimethod dispatch.

@@

The final example combines everything we will cover for the rest of the paper:
multimethod dispatch, reflection resolution via type hints, Java method
and constructor calls, conditional and exceptional flow reasoning
and use of HMaps. 
\begin{Code}
\begin{exmp}
\inputminted[firstline=6,lastline=20]{clojure}{code/demo/src/demo/eg8.clj}
\end{exmp}
\end{Code}
\clj{PayLoad} is either a HMap containing a file
or a string. We dispatch on \clj{:p} to distinguish the two cases---for example on \clj{:F}
we know the \clj{:file} is a file.
The body of the first method uses type hints to resolve reflection
and conditional control flow to prove null-pointer exceptions are impossible.
The second method is similar except it uses exceptional control flow.

>>>>

We introduce Clojure's syntax.
Parenthesis indicate \emph{applications}, square brackets
are syntax for literal \emph{vectors} and curly braces delimit literal
\emph{hash-maps}.
\clj{nil} is exactly Java's \java{null} and is one of two logically
false values (the other is \clj{false}).
\emph{Symbols} begin with an alphabetic character.
A symbol prefixed with a colon like \clj{:a} is an interned \emph{keyword}.
The \clj{get} function takes a hash-map and a key returns the value
associated with the key in the map, or returns \clj{nil} if there is none.
Keywords are functions that take a map and applies \clj{get} to the map
and itself.
\clj{inc} and \clj{dec} increment and decrement numbers respectively.
\clj{assoc} and \clj{dissoc} associate and dissociate entries
from immutable maps.
The \clj{=} function returns true if all its arguments are pointer equal.
Everything is an instance of some class
except for \clj{nil}. Predicates like \clj{number?} are equivalent to
a Java \java{instanceof} test, in this case on \clj{java.lang.Number}.

Clojure uses \emph{namespaces}, declared with the
\clj{ns} form, to manage top-level \emph{vars} and
class imports.
Example~\ref{example:conditionalflow}
is a regular Clojure file compiled with
the Clojure compiler using \clj{:require} to declare a
runtime dependency on Typed Clojure's core namespace
\clj{clojure.core.typed}.
Programmers import typed versions of certain constructs as needed, like
a \clj{fn} variant that supports annotations,
and use the provided \clj{check-ns} function to type check the current namespace
when convenient.

\emph{Map destructuring} is binding position syntax for pattern matching.
Parameters are replaced with a map containing entries of symbols to keywords that bind the symbols
to lookups on that keyword with the current argument---the \clj{:as} option
can follow which binds a symbol to the current argument. For example
$$
\clj{(fn [{^File x :x, :as m}] ...)}
$$ 
expands to a \clj{let} that binds \clj{m} to the argument and 
\clj{^File x} to \clj{(:x m)}.

@@

@@

Along with conditional control flow,
Clojure programmers rely on \emph{exceptions}
to assert type-related invariants, such as preventing
null-pointer exceptions.

@@

To check this example,
occurrence typing 
automatically
assumes
\clj{x} is a number when checking the second \clj{do} subexpression
based on the first subexpression.
\footnote{We omit \clj{ns} forms for the rest of the examples, see <this> github repo for full working examples}
We model this formally (section~\ref{sec:doformal}) and prove
null-pointer exceptions are impossible in typed code (section~\ref{sec:metatheory}).

@@

\begin{exmp}
\inputminted[firstline=6,lastline=13]{clojure}{code/demo/src/demo/hmap.clj}
\label{example:decleaf}
\end{exmp}

Here \clj{defalias} defines a type abbreviation called \clj{Expr}
that describes the structure of a recursively-defined AST as a union of HMaps.
\clj{an-exp} is a function verified to return an \clj{Expr}.

The next example is extracted from a Typed Clojure production system at CircleCI
(section~\ref{sec:casestudy} presents a case study).

\begin{exmp}
\inputminted[firstline=10,lastline=22]{clojure}{code/demo/src/demo/key.clj}
\end{exmp}

\clj{enc-keypair} takes an unencrypted keypair and returns an encrypted keypair by
dissociating the raw \clj{:private-key} entry and associating an encrypted private key
as \clj{:enc-private-key}.
Since \clj{EncKeyPair} is \clj{:complete?}, Typed Clojure enforces the return type
does not contain an entry \clj{:private-key}, and would complain if the \clj{dissoc}
operation forgot to remove it.

\begin{exmp}
\inputminted[firstline=10,lastline=21]{clojure}{code/demo/src/demo/key2.clj}
\end{exmp}

Next, 
HMaps interact with occurrence typing in interesting ways.

\begin{exmp}
\inputminted[firstline=15,lastline=27]{clojure}{code/demo/src/demo/hmap.clj}
\end{exmp}

@@

Clojure supports interoperability with Java, including the ability to
call constructors, methods and access fields.

@@

Unlike Java, Clojure is dynamically typed. We have no type information about \clj{f}
but we still need to pick a method to call. The Clojure compiler delegates
the choice to runtime using \emph{Java reflection}.
Unfortunately reflection is slow and unpredictable, so Clojure supports \emph{type hints}
to help eliminate it where possible,

@@

The Clojure compiler uses the type hint
to statically resolve 
the method call to the
\mintinline{java}{public String getParent()}
method of \clj{java.io.File}.
The method is unambiguous but type checking fails---Typed Clojure
considers \clj{f} to be of type \clj{Any}, which is unsafe 
to use as the target of even a resolved method.
We annotate the function parameter with type \clj{File}.

\begin{minted}{clj}
  (fn [f :- File] (.getParent f)) ;; type error
\end{minted}

Typed Clojure disallows reflection in typed code
so we must add back the type hint to obtain a well-typed
expression.

@@

The type hinting system and Typed Clojure's static type checking
are separate, the latter predating the former by several years.
The interaction between them is often not as obvious, for example type hints
do not need to take \clj{nil} into account, while Typed Clojure guarantees
null-pointer exceptions are impossible.

\begin{exmp}
\inputminted[firstline=18,lastline=19]{clojure}{code/demo/src/demo/parent3.clj}
\end{exmp}

@@

\begin{exmp}
\inputminted[firstline=6,lastline=8]{clojure}{code/demo/src/demo/parent3.clj}
\end{exmp}

@@

\begin{exmp}
\inputminted[firstline=15,lastline=16]{clojure}{code/demo/src/demo/parent3.clj}
\end{exmp}

By default Typed Clojure conservatively assumes method and constructor arguments to be \emph{non-nullable},
but can be configured globally for particular positions if needed.

@@

A multimethod in Clojure is a function that contains a dispatch
function and methods. Multimethods are created with {\clj{defmulti}}.
\inputminted[firstline=5,lastline=7]{clojure}{code/demo/src/demo/rep.clj}
The multimethod \clj{path} has type \clj{[Any -> String]}, an initially empty \emph{dispatch table}
and \emph{dispatch function} \clj{class}, a function that returns the argument's class or \clj{nil} if none.
There are currently no methods so \clj{path} will fail at runtime.
\begin{minted}{clojure}
  (path "dir/a") ;=> IllegalArgumentException ...
\end{minted}

We can use {\clj{defmethod}} to add a method to \clj{path}.
\inputminted[firstline=8,lastline=8]{clojure}{code/demo/src/demo/rep.clj}
Now \clj{path}'s dispatch table contains a method mapping
the \emph{dispatch value} \clj{String} to the function
\clj{(fn [x] x)}. The call \clj{(path "dir/a")}
now dispatches to the method because
\clj{(isa? (class "dir/a") String)}
is true, and finally returns \clj{((fn [x] x) "dir/a")}.
\clj{isa?} is a subclassing check when provided with classes,
otherwise it is an equality check.
\begin{minted}{clojure}
  (isa? Keyword Object) ;=> true
  (isa? :a :a) ;=> true
\end{minted}

\begin{exmp}
\inputminted[firstline=5,lastline=11]{clojure}{code/demo/src/demo/rep.clj}
\label{example:rep}
\end{exmp}

Typed Clojure does not statically ensure multimethod calls dispatch successfully---\clj{(path :a)} 
type checks but throws a runtime error.

The flexibility of \clj{isa?} is key to the generality of multimethods. 
We can dispatch on the \clj{:op} key 
of our HMap AST \clj{Expr}.
Keywords are functions that look themselves up in their argument, so \clj{:op}
is our dispatch function.

\begin{exmp}
\inputminted[firstline=5,lastline=25]{clojure}{code/demo/src/demo/eg5.clj}
\end{exmp}

\clj{isa?} is special with vectors---vectors of the
same length recursively call \clj{isa?} on the elements pairwise.
\begin{minted}{clojure}
  (isa? [Keyword Keyword] [Object Object]) ;=> true
\end{minted}

Now we can simulate multiple dispatch by dispatching on
a vector containing the class of both arguments.

\begin{exmp}
\inputminted[firstline=6,lastline=22]{clojure}{code/demo/src/demo/eg7.clj}
\end{exmp}

We call three different \clj{File} constructors, each known at compile-time
via type hints.
No extra annotations are needed to follow type-directed control flow
in multimethod dispatch.

@@

The final example combines everything we will cover for the rest of the paper:
multimethod dispatch, reflection resolution via type hints, Java method
and constructor calls, conditional and exceptional flow reasoning
and use of HMaps.

\begin{exmp}
\inputminted[firstline=6,lastline=20]{clojure}{code/demo/src/demo/eg8.clj}
\end{exmp}

\clj{PayLoad} is either a HMap containing a file
or a string. We dispatch on \clj{:p} to distinguish the two cases---for example on \clj{:F}
we know the \clj{:file} is a file.
The body of the first method uses type hints to resolve reflection
and conditional control flow to prove null-pointer exceptions are impossible.
The second method is similar except it uses exceptional control flow.
